Correlation among smart grid actors

### 2015-04-29 Article.tex
update

Designing stable power systems is a classical engineering challenge since blackout events can have catastrophic consequences. Nevertheless, stability is a general concept and can be studied from several points of view. A recent approach introduced by complex systems theorists consists in abstracting the power grid as a graph where nodes represent loads, generators, or transformers, while edges stand for electrical lines. As graphs, power grids display some characteristics like clustering coefficients, degree distributions, or density for instance. These metrics enable scientists to get a better understanding of the structure of such complex systems. It appears that power grids are particular graphs compared to well-known large networks like the internet, the world wide web, or social networks. That is, they exhibit different characteritics, suggesting that they tend to behave differently. A common hypothesis for such a difference is that power grids are completely designed by a very small number of entities (electricity operators mainly), whereas other pre-cited graphs results from the interactions of billions of independant entities. Therefore, meaningfull properties like power law degree distributions, prerential attachment, or hub oriented attacks do not seem to apply with power grids.

@@

The constraints necessary for the electrical grid to remain in a stable state obiously forbid any anarchic system where prosumers inject power whenever and however they want. A solution for organizing such a system consists in a market environement, where participating entities announce a production capacity for an upcoming period of time. These entities thus commit to injecting exactly at any time the contracted amount of power under financial penalities if they fail. Such a market system necessitates obviously some communication architecture to be viable. Since the number of prosumers in a system may be large, it seems unrealistic to have a flat centralized topology where a single agent coordinates all the others individually. Moreover, taking into account that the production component of these prosumers is supposed to come mainly from renewables, the over-producing state of the prosumers, and therefore the power injected in the grid, might be rather unstable. Imagine indeed how strong the impact of a simple cloud on a single agent production will be.

It has been suggested that energy diversification, geographical expansion, as well as using storage devices are envisageable techniques for  stabilizing the production. Considering virtual coalitions of prosumers as potential providers, and allowing these aggregations to enter the market, could potentially solve these problems. Indeed, the system operator would only have to communicate with a relatively small number of aggregators, that in turn will use some internal communication protocol to manage the involved agents. Moreover, if the aggregation has been done properly, one expect a more stable and predictable energy production for the coalition than for a single agent. This diversification idea is ideed a central topic of the present paper : given N prosumers, what coalitions should be formed so that the compromise between expected production and variability is optimized ?

@@

This problem exhibits similarities with our objective of forming stable coalitions since our expected return can be formulated as the expected production and the risk as the variance of the production. However, one of the assumptions in the portfolio theory is that returns are normaly distributed random variables (or at least that the joint distribution is elliptic). This is a strong assumption which reveals unrealistic in most cases for our purposes when looking at real data.

@@

An efficient portfolio of assets and a clustering of assets based on the correlations are obviously two very distinct things. And none of them is exactly what we seek. We will see in section X, how we combined both theories in order to form sufficiently producing coalitions with an acceptable risk level.

@@

An essential component of the smart grid is the smart meter which makes the interface between the end user and the rest of the system. Smart meters coupled with sensors measure quantities of interest like instantaneous consumption, receive informations from the grid (electricity price for instance), and take actions accordingly (demand side management program). Smart meters are currently and gradually deployed, and will probably provide interesting datasets to work on. Unfortunately, at the time this paper was written, production and consumption data for prosumers over a large region were not yet available to our knowledge. Some interesting experiments are notwithstanding being conducted and data are progressively made public (see for instance the ISSDA experiment in Ireland).

We denote by $ P_{i}(t) $ the instantaneous extra-production of agent i at time t. That is, $ P_{i}(t) = P_{i}^{P}(t) - P_{i}^{D}(t) $, where $ P_{i}^{P}(t) $ represents the total production of agent i at time t and $ P_{i}^{D}(t) $ its consumption at time t. In other words, $ P_{i}(t) $ represents the instantaneous surplus of power that agent i is willing to sell at time t. As explained above, since large datasets containing this quantity over time are not yet available, we simulated these traces by considering separately $ P_{i}^{P} $ and $ P_{i}^{D} $.

@@

Weather quantities like wind speed or solar irradiance appear thus as alternative data for generating the $ P_{i} $ series. Fortunately, these kind of data are easier to find, and since the developement of small personal weather stations, their geographical granularity keeps increasing.  A quantity like wind speed depends indeed both on time and location. In the following, time is discretized into slots and space into zones. A zone is simply a portion of the considered space for which we have weather data. Therefore, if prosumers i and j are positioned on the same zone, they are exposed to the same weather. Adding some noise can easily be done though not considered in this paper. The process for generating the $ P_{i} $ series is pictured in the first block of the process diagram (see figure \ref{fig:process}).

@@

$ \rho_{ij} $ being the Pearson's correlation coefficient between $ P_{i} $ and $ P_{j} $. If the coalition $ \{i,j\}$ proposes a contract value $ P_{S}^{CRCT} $, all instants where $ \{i,j\}$ will produce less than $ P_{S}^{CRCT} $ is critical. Indeed, in this situation, $ \{i,j\}$ will either have to discharge batteries to keep up with its contract, or pay penalities to the grid. The probability that $ \{i,j\}$ is underproducing compared to the contract : $ Pr[P_{i,j} \leq P^{CRCT}] $ is thus an important indicator of the coalition's credibility on the market. A well-known result for normal distributions is that the cumulative distribution function can be written as :
\begin{equation}
Pr[P_{ij} \leq P^{CRCT}] = \dfrac{1}{2} \left[ 1+ erf \left( \dfrac{P^{CRCT} - \mu_{ij}}{\sigma_{ij}\sqrt{2}} \right) \right] 
\end{equation}

The amount of risks a given coalition is willing to take depends on a lot of things, among wich its capacity to compensate for under-producing (using batteries, backup generators...). Selecting the right contract value appears thus as an interesting problem on its own that we plan to investigate in future works. In order to keep the present paper in a reasonable length, we simplify the contract value selection problem a little bit by giving some responsabilities to a third party named the grid operator. The role of the grid operator is to constrain the market entry with an absolute low bound on the contract value and a upper limit relative to the coalition. In other words, the grid operator restrict the market to coalitions able to propose both sufficiently high and sufficiently credible contract values. More formally, let $ \phi \in [0,1] $ be a threshold fixed by the grid operator as a maximum value for the probability of under-producing. The highest contract value that a coalition can propose is thus $ P_{S}^{CRCT \star} $ such that $ Pr[P_{ij} \leq P_{S}^{CRCT \star}] = \phi $. In the gaussian example, this translates by coalition $ \{i,j\}$ announcing :

@@

\begin{figure}
\includegraphics[scale=.6]{gaussian_P_star.pdf}
\caption{$ P_{S}^{CRCT \star} $ depending on reliability parameter $ \phi $ for gaussian distributions (see equation \ref{eq:P_star}). Blue curve with triangles stands for a coalition S with an expected production of 5 units and a standard deviation of 0.5. Under a grid policy of $ \phi = 0.1 $, it is able to announce a contract value of $ P_{S}^{CRCT \star} = 4.36 $. The same coalition in term of expected production ($\mu = 5 $), but with a higher variance ($ \sigma = 1.5 $, cyan curve with circles) can only afford a smaller contract value of $ P_{S}^{CRCT \star} = 3.07 $. The red curve with diamonds stands for a coalition with a higher expected production $(\mu = 7)$, but with a very high unpredictability ($ \sigma = 5 $). For low values of $ \phi $, this coalition is thus heavily penalized and can only afford a contract of $ 0.59 $ units. Under grid policy $ (\phi = 0.1, P^{MIN} = 2) $, this last coalition is thus not allowed to enter the market (red dot below the horrizontal dashed line)	. }
\label{fig:Gaussian}
\end{figure}

This is the best contract value that the coalition S can afford giving the stability policy $\phi$ of the grid operator, and it is computable by any coalition. Figure \ref{fig:Gaussian} shows how $ P_{S}^{CRCT \star} $ evolves according to the reliability parameter $ \phi $. For illustration, the range of $ \phi $ values is shown from 0 to 1, but in practice, only small values of $ \phi $ really make sense : $ \phi = 1 $ for instance means that coalitions can announce absolutely anything since the probability of producing less than any contract value is necessarily less than one by trivial definition of a probability. As visible on figure \ref{fig:Gaussian}, coalitions with high expected production but presenting a high unpredictability are penalized and can only afford small contracts.

@@

\begin{equation}
\left\{ \begin{array}{lll}
			Pr[P_{ij} \leq P^{CRCT}] \leq \phi \\
			P^{CRCT} \geq P^{MIN}
\end{array} \right.
\end{equation}

@@

Since $ \lambda $ appears, in this simple model, just a multiplicative constant, we consider for simplicity $ \lambda = 1 $ in the following. The utility of S is now the amount of energy S plans to inject during its contract period. Since we focused on power quantities from the begining of the present paper, and because the power injected by S in the grid is supposed to be constant over time, we simplify further the utility to its contract value : $ \mathcal{U}(S) = P_{S}^{CRCT} $. Although concise, this formulation suffers a major drawback. It is indeed not a concave function of the coalition length, meaning that coalitions can grow as large as the number of agents allows it, without any counterparty.

Such a model, that virtually allows infinitely large coalitions, is in practice not realistic. There are indeed costs (communication costs for instance) that increase with the coalitions sizes. We take this observation into account by rescaling the utility of a coalition by its size in term of number of agents :

\begin{equation}
\mathcal{U}(S) = \left\{ \begin{array}{lll}
							\dfrac{1}{|S|^{\alpha}} \dfrac{ P_{S}^{CRCT} }{P_{S}^{MAX}},\ if\ S\ is\ valid, \\
							0,\ if\ S\ is\ not\ valid
						 \end{array}
				  \right.
\end{equation}

Where parameter $ \alpha $ controls to what extent the size of the coalition impacts its utility, and $ P_{S}^{MAX} $ is a normalizing factor. $ P_{S}^{MAX} $ is the maximum production possible, which only occurs when all consumptions are null and all generators produce their maximum power.

@@

\begin{equation}
\delta_{S}(i) \geq 0 \Leftrightarrow P_{S+\{i\}}^{CRCT} \geq P_{S}^{CRCT} \left( \dfrac{|S|+1}{|S|} \right)^{\alpha} \dfrac{P_{S + \{i\}}^{MAX}}{P_{S}^{MAX}}
\label{eq:marginal_benefit}
\end{equation}

In order to understand how this utility function behaves and how it relates to the $ \alpha $ parameter, we use again the gaussian case (as in section \ref{sec:notations}). Since considering all the possible interactions between agents is analytically intractable, we use here a mean approximation. Any quantity x that varies over the agent set is thus simplified in its mean value $ \bar{x} $. In the following $ \bar{N} $ denotes the mean number of agents in a coalition. If we wish that $ \mathcal{U}(S) $ tends to form coalitions of size approximately of the order of $ \bar{N} $, then :

\begin{equation}
\left[ \dfrac{\partial{ U}}{ \partial{|S|}} \right]_{|S| = \bar{N}} = 0
\end{equation}

Solving this equation for $ \alpha $ in the mean approximation for the gaussian case yields :
\footnotesize
\begin{equation}
\alpha^{\star}_{\bar{N}} = \dfrac{0.7 \bar{\sigma}(\bar{\rho}-1)erf^{-1}(2 \phi - 1)}{\bar{\mu}\sqrt{\bar{N}(\bar{\rho}\bar{N}-\bar{\rho}+1)}+1.4 \bar{\sigma} erf^{-1} (2 \phi -1 ) (\bar{\rho}\bar{N}-\bar{\rho} + 1)} 
\label{eq:alpha_star}
\end{equation}
\normalsize

@@

As can be pointed out, the purpose of $ \mathcal{U} $ is not a study of coalitions stability against player defection, wich game theory provides a lot of tools for. This is indeed a problem on its own. The redistribution of a coalition's utility in terms of individual payoffs will therefore not be considered directly in this paper. $ \mathcal{U} $ can be rather interpreted as a measure of how good a given coalition is according to our criteria since it favors small coalitions with a good production to risk ratio.

@@

The previous section explained that coalitions with small variances in their production probability distributions are more likely to afford high contracts. Furthermore, the gaussian approximation highlighted that the correlation structure of the agents plays an important role in finding stable coalitions. Usually, this correlation structure is formalized with a covariance matrix or a correlation matrix that contains all the correlation coefficients between the agents : $ M = (\rho_{ij})_{\forall i,j \in \mathcal{A}^{2}}$.

@@

Selecting the right filter $ \epsilon $ is an important point since it affects the correlation structure which affects in turn the formation of the coalitions. Unfortunately, there seems to be no clear consensus in the litterature on how to select such a threshold. In our situation, we want to generate $ N_{COAL} $ coalitions from cliques in the graph. We need therefore at least $ N_{COAL} $ cliques of a given size to start. Besides, since we consider coalitions as disjoint sets, the starting cliques should be non overlapping. We thus select the optimal threshold as :

@@

Where $ G_{2}^{\epsilon} $ is the de-correlation graph $ G_{2} $ filtered by $ \epsilon $, and $ \Theta_{k}(G) $ is the set of non overlapping cliques of size k in a given graph G. In other words we select $ \epsilon $ as the smallest threshold posible such that the filtered de-correlation graph contains at least $ N_{COAL} $ non overlapping cliques of size k. The existence of $ \epsilon^{\star} $  as defined in equation  \ref{epsilon_star} is not guaranteed. The users has indeed to provide consistent values of $ N_{COAL} $ or k compared to the size of the agent population $ \mathcal{A} $.

Correlation can be seen as cosine of angles in $ L^{2} $, hence even if there is no strict transivity relation for correlation, there is, to a certain extent, some partial notion of it. More precisely, if a, b, and c are three items such that $ \rho_{ab} > \delta $ and $ \rho_{bc} > \delta $, then we know, by the cosine addition formula\footnote{ $ cos(a+b) = cos(a)cos(b) - sin(a)sin(b) $ }, that $ \rho_{ac} > 2 \delta^{2} - 1 $. That is, if a and b are strongly correlated ($\rho_{ab} > 0.9 $) and b and c are also strongly correlated ($\rho_{bc} > 0.9 $), then there is a high probability for a and c of being strongly correlated ($\rho_{ac} > 0.62 $). This is one of the reasons why searching for clusters in correlation graphs, that is, clustering according to the correlation, makes sense.

Nevertheless de-correlation seems like a more complex concept than correlation in the sense that there is not even a partial notion of transitivity when it comes to it. As expected, the clustering coefficients of $ G_{1} $ is much higher than the one of $ G_{2} $. This can be seen as another formulation of Onnela's study on the structural roles of weak and strong links on correlation graphs. Strong links, accounting for strong correlation relationships, are responsible for the clustering, while weak links provide the connectivity between clusters. Searching for clusters in $ G_{2} $ and hoping that this strategy will provide a nice coalition structure of internally uncorrelated coalitions seems thus pointless.

Consider now a clique in  $ G_{2} $, which is a complete subgraph of $ G_{2} $. This is indeed a structure of interest for our purpose. Since there is a link for every pairs of nodes, we know, by construction, that a clique has a mean correlation and a maximum correlation less than $ \epsilon $. More formally, let

\begin{equation}
\left\{ \begin{array}{lll}
			\bar{\rho}_{S} = \dfrac{2}{|S|(|S|-1)} \sum_{i \in S} \sum_{j \in S, j>i} \rho_{ij}, \\ 
			\rho_{S}^{>} = MAX_{(i,j) \in S^{2}}(\rho_{ij})
\end{array} \right.			
\end{equation}

be the mean correlation of coalition S and the maximum correlation value in S. If S is a clique in $ G_{2} $ and $ \epsilon \in [0,1] $ is the threshold used for filtering $ G_{2} $, then $ \bar{\rho}_{S} \leq 2/(|S|(|S|-1)) \sum_{i \in S} \sum_{j \in S, j>i} \epsilon $, that is, $ \bar{\rho}_{S} \leq \epsilon $, and $ \rho_{S}^{>} \leq \epsilon $.

@@

The algorithm presented in the previous section is supposed to generate a given number of coalitions that have good utilities, and therefore, high contract values for relatively small sizes. Nevertheless, as it comprises mainly of a greedy optimization based on local improvements, the probability that the algorithm finds the optimal coalitions set is very low. Actually, it is not obvious that a strict optimum exists at all. Besides, there is no, to our knowledge, state of the art algorithm that aggregate uncorrelated series in an optimum way.

In order to have an idea about the algorithm's quality, we compare its results with :
\begin{itemize}
\item \textbf{Random :} This algorithm is basically a random search over the coalition structures space. It analyses a given number of structures and returns the best that it has encountered so far. See algorithm \ref{alg:algo2}.
\item \textbf{Correlated :} This is the complete opposite of our algorithm. It basically uses the correlation graph $ G_{1} $ and performs community detection on it. The resulting coalitions have thus very high internal correlations. We thus expect this algorithm to perform very bad compared to the others. See algorithm \ref{alg:algo3}.
\item \textbf{K-means :} This is simply a K-means clustering of the agents $ P_{i} $'s series. The goal of presenting this algorithm is to show that K-means do not consider correlations explicitely and is therefore not well suited for our purpose.
\end{itemize}

The coalitions formed according to these algorithms will be compared using two criteria : the normalized global utility of the coalitions formed $ \bar{\mathcal{U}}(CS) \in [0,1] $, and the percentage of valid coalitions that satisfy the grid conditions $ \tau(CS) \in [0,1] $ :
\begin{equation}
	\bar{\mathcal{U}}(CS) = \dfrac{1}{N_{COAL}} \sum_{S \in CS} \mathcal{U}(S)
\end{equation}
\begin{equation}
	\tau(CS) = \dfrac{|\{ S \in CS,\ s.t\ \mathcal{U}(S) > 0 \}|}{N_{COAL}}
\end{equation}

Before running the algorithm, we need to calibrate the utility function by choosing the $ \alpha $ parameter. In this paper, neither the communication network nor the electrical grid are explicitely considered. Thus, we do not have any technical constraints on coalition sizes even if we designed the utitility such that these kind of constraints could be taken into account. We select the desired size as being $ \lfloor N/N_{COAL} \rfloor $ (where $ \lfloor.\rfloor $ means floor). Figure \ref{fig:real_utility2} shows how the mean utility of a coalition evolves with its size when the optimum size is set to 40 agents. Using equation \ref{eq:alpha_star} to select $ \alpha $ based on the mean quantities seems to give acceptable results for the utility function.

@@

>>>>

Designing stable power systems is a classical engineering challenge since blackout events can have catastrophic consequences. A recent approach introduced by complex systems theorists consists in abstracting the power grid as a graph where nodes represent loads, generators, or transformers, while edges stand for electrical lines. As graphs, power grids display some characteristics like clustering coefficients, degree distributions, or density for instance. These metrics enable scientists to get a better understanding of the structure of such complex systems. It appears that power grids are particular graphs compared to well-known large networks like the internet, the world wide web, or social networks. That is, they exhibit different characteritics, suggesting that they may behave differently. A common hypothesis for such a difference is that power grids are completely designed by a very small number of entities (electricity operators mainly), whereas other pre-cited graphs results from the interactions of billions of independant entities. Therefore, properties like power law degree distributions, prerential attachment, or hub oriented attacks do not necessarily apply with power grids.

@@

The constraints necessary for the electrical grid to remain in a stable state obiously forbid any anarchic system where prosumers inject power whenever and however they want. An envisaged solution for organizing such a system consists in a market environement, where participating entities announce a production capacity for an upcoming period of time. These entities thus commit to injecting exactly at any time the contracted amount of power under financial penalities if they fail. Such a market system necessitates obviously some communication architecture to be viable. Since the number of prosumers in a system may be large, it seems unrealistic to have a flat centralized topology where a single agent coordinates all the others individually. Some articles suggest that a fully decentralized architecture will probably be too complicated to achieve, and that some kind of hierarchical hybrid topology seems more likely to be viable.

Moreover, taking into account that the production component of a prosumer is supposed to come mainly from renewables, its over-producing state, and therefore the power he injects in the grid, might be rather unstable.

It has been suggested that energy diversification, geographical expansion, as well as using storage devices are envisageable techniques for  stabilizing the production. Considering virtual coalitions of prosumers as potential providers, and allowing these aggregations to enter the market, could potentially solve some of these problems. Indeed, the system operator would only have to communicate with a relatively small number of aggregators, that in turn will use some internal communication protocol to manage the involved agents. Moreover, if the aggregation has been done properly, one should expect a more stable and predictable energy production for the coalition than for a single agent. This diversification idea is ideed a central topic of the present paper : given N prosumers, what coalitions should be formed so that the compromise between expected production and variability is optimized ?

@@

This problem exhibits similarities with our objective of forming stable coalitions since our expected return can be formulated as the expected production and the risk / volatility as the variance of the production. However, one of the assumptions in the portfolio theory is that returns are normaly distributed random variables (or at least that the joint distribution is elliptic). This is a strong assumption which reveals unrealistic in most cases for our purposes when looking at real data.

@@

An efficient portfolio of assets and a clustering of assets based on the correlations are obviously two very distinct things. And none of them is exactly what we seek. We will see in section X, how we combined both theories in order to form sufficiently producing coalitions with an acceptable volatility level.

@@

An essential component of the smart grid is the smart meter which makes the interface between the end user and the rest of the system. Smart meters coupled with sensors measure quantities of interest (like instantaneous consumption), receive informations from the grid (electricity price for instance), and take actions accordingly (demand side management program). Smart meters are currently and gradually deployed, and will probably provide interesting datasets to work on. Unfortunately, at the time this paper was written, production and consumption data for prosumers over a large region were not yet available to our knowledge. Some interesting experiments are notwithstanding being conducted and data are progressively made public (see for instance the ISSDA experiment in Ireland).

In the following, we denote by $ P_{i}(t) $ the instantaneous extra-production of agent i at time t. That is, $ P_{i}(t) = P_{i}^{P}(t) - P_{i}^{D}(t) $, where $ P_{i}^{P}(t) $ represents the total production of agent i at time t and $ P_{i}^{D}(t) $ its consumption at time t. In other words, $ P_{i}(t) $ represents the instantaneous surplus of power that agent i is willing to sell at time t. As explained above, since large datasets containing this quantity over time are not yet available, we simulated these traces by considering separately $ P_{i}^{P} $ and $ P_{i}^{D} $.

@@

Weather quantities like wind speed or solar irradiance appear thus as alternative data for generating the $ P_{i} $ series. Fortunately, these kind of data are easier to find, and since the developement of small personal weather stations, their geographical granularity keeps increasing.  A quantity like wind speed depends both on time and location. In the following, time is discretized into slots and space into zones. A zone is simply a portion of the considered space for which we have weather data. Therefore, if prosumers i and j are positioned on the same zone, they are exposed to the same weather. Adding some noise can easily be done though not considered in this paper. The process for generating the $ P_{i} $ series is pictured in the first block of the process diagram (see figure \ref{fig:process}).

@@

$ \rho_{ij} $ being the Pearson's correlation coefficient between $ P_{i} $ and $ P_{j} $. If the coalition $ \{i,j\}$ proposes a contract value $ P_{S}^{CRCT} $, all instants where $ \{i,j\}$ will produce less than $ P_{S}^{CRCT} $ is critical. Indeed, in this situation, $ \{i,j\}$ will either have to discharge batteries to keep up with its contract, or pay penalities to the grid. The probability that $ \{i,j\}$ is underproducing compared to the contract : $ Pr[P_{i,j} \leq P^{CRCT}] $ is thus an important indicator of the coalition's quality. A well-known result for normal distributions is that the cumulative distribution function can be written as :
\begin{equation}
Pr[P_{ij} \leq P^{CRCT}] = \dfrac{1}{2} \left[ 1+ erf \left( \dfrac{P^{CRCT} - \mu_{ij}}{\sigma_{ij}\sqrt{2}} \right) \right] 
\end{equation}

Where $ erf $ is the error function : $ erf(x) = \dfrac{2}{\sqrt{\pi}}\int_{0}^{x} e^{-t^{2}} dt $.

The amount of risks a given coalition is willing to take depends on a lot of things, among wich its capacity to compensate for under-producing (using batteries, backup generators...). Selecting the right contract value appears thus as an interesting problem on its own that we plan to investigate in future works. In order to keep the present paper in a reasonable length, we simplify the contract value selection problem a little bit by giving some responsabilities to a third party named the grid operator. The role of the grid operator is to constrain the market entry to coalitions able to propose both sufficiently high and sufficiently credible contract values. More formally, let $ \phi \in [0,1] $ be a threshold fixed by the grid operator as a maximum value for the probability of under-producing. The highest contract value that a coalition can propose is thus $ P_{S}^{CRCT \star} $ such that $ Pr[P_{ij} \leq P_{S}^{CRCT \star}] = \phi $. In the gaussian example, this translates by coalition $ \{i,j\}$ announcing :

@@

\begin{figure}
\includegraphics[scale=.6]{gaussian_P_star.pdf}
\caption{$ P_{S}^{CRCT \star} $ depending on reliability parameter $ \phi $ for gaussian distributions (see equation \ref{eq:P_star}). Blue curve with triangles stands for a coalition S with an expected production of 5 units and a standard deviation of 0.5. Under a grid policy of $ \phi = 0.1 $, it is able to announce a contract value of $ P_{S}^{CRCT \star} = 4.36 $. The same coalition in term of expected production ($\mu = 5 $), but with a higher variance ($ \sigma = 1.5 $, cyan curve with circles) can only afford a smaller contract value of $ P_{S}^{CRCT \star} = 3.07 $. The red curve with diamonds stands for a coalition with a higher expected production $(\mu = 7)$, but with a very high unpredictability ($ \sigma = 5 $). For low values of $ \phi $, this coalition is thus heavily penalized and can only afford a contract of $ 0.59 $ units. Under grid policy $ (\phi = 0.1, P^{MIN} = 2) $, this last coalition is thus not allowed to enter the market (red dot below the horrizontal dashed line). }
\label{fig:Gaussian}
\end{figure}

This is the best contract value that the coalition S can afford giving the stability policy $\phi$ of the grid operator. Figure \ref{fig:Gaussian} shows how $ P_{S}^{CRCT \star} $ evolves according to the reliability parameter $ \phi $. For illustration, the range of $ \phi $ values is shown from 0 to 1, but in practice, only small values of $ \phi $ really make sense : $ \phi = 1 $ for instance means that coalitions can announce absolutely anything since the probability of producing less than any contract value is necessarily less than one by trivial definition of a probability. As visible on figure \ref{fig:Gaussian}, coalitions with high expected production but presenting a high unpredictability are penalized and can only afford small contracts.

@@

\begin{equation}
\left\{ \begin{array}{lll}
			Pr[P_{ij} \leq P_{S}^{CRCT}] \leq \phi \\
			P_{S}^{CRCT} \geq P^{MIN}
\end{array} \right.
\end{equation}

@@

Since $ \lambda $ appears, in this simple model, just a multiplicative constant, we consider for simplicity $ \lambda = 1 $ in the following. The utility of S is now the amount of energy S plans to inject during its contract period. Since we focused on power quantities from the begining of the present paper, and because the power injected by S in the grid is supposed to be constant over time, we simplify further the utility to its contract value : $ \mathcal{U}(S) = P_{S}^{CRCT} $. Although concise, this formulation suffers a major drawback. It is indeed not a concave function, meaning that coalitions can grow as large as the number of agents allows it, without any counterparty.

Such a model, that virtually allows infinitely large coalitions, is in practice not realistic. There are indeed costs (communication costs for instance) that increase with the coalitions sizes. We take this observation into account by rescaling the utility of a coalition S by its size in term of number of agents ($|S|$):

\begin{equation}
\mathcal{U}(S) = \left\{ \begin{array}{lll}
							\dfrac{1}{|S|^{\alpha}} \dfrac{ P_{S}^{CRCT} }{P^{MAX}},\ if\ S\ is\ valid, \\
							0,\ if\ S\ is\ not\ valid
						 \end{array}
				  \right.
\end{equation}

Where parameter $ \alpha $ controls to what extent the size of the coalition impacts its utility, and $ P^{MAX} $ is a normalizing factor. $ P^{MAX} $ can be seen as the maximum production which can be injected in the grid.

@@

\begin{equation}
\delta_{S}(i) \geq 0 \Leftrightarrow P_{S+\{i\}}^{CRCT} \geq P_{S}^{CRCT} \left( \dfrac{|S|+1}{|S|} \right)^{\alpha}
\label{eq:marginal_benefit}
\end{equation}

If $ \alpha $ is set to zero, agents are added as long as they increase the contract value of the coalition. If $ \alpha $ is greater than zero, additional agents have to increase the contract value by some factor. Since tunning the utility function with $ \alpha $ is not necessarily convenient, we relate $ \alpha $ to the mean sizes of the coalitions $ \bar{N} $. If we wish that $ \mathcal{U}(S) $ tends to form coalitions of size approximately of the order of $ \bar{N} $, then :

\begin{equation}
\left[ \dfrac{\partial{ U}}{ \partial{|S|}} \right]_{|S| = \bar{N}} = 0
\label{eq:derivative}
\end{equation}

In order to get an estimator for $ \alpha $, we solve equation \ref{eq:derivative} in a gaussian case (as in section \ref{sec:notations}). Furthermore, since considering all the possible interactions between agents is analytically intractable, we use here a mean approximation. Any quantity x that varies over the agent set is thus simplified in its mean value $ \bar{x} $. Solving equation \ref{eq:derivative} for $ \alpha $ in these conditions yields :

\footnotesize
\begin{equation}
\alpha^{\star}_{\bar{N}} = \dfrac{0.7 \bar{\sigma}(\bar{\rho}-1)erf^{-1}(2 \phi - 1)}{\bar{\mu}\sqrt{\bar{N}(\bar{\rho}\bar{N}-\bar{\rho}+1)}+1.4 \bar{\sigma} erf^{-1} (2 \phi -1 ) (\bar{\rho}\bar{N}-\bar{\rho} + 1)} 
\label{eq:alpha_star}
\end{equation}
\normalsize

@@

As can be pointed out, the purpose of $ \mathcal{U} $ is not a study of coalitions stability against player defection, wich game theory provides a lot of tools for. This is indeed a problem on its own. The redistribution of a coalition's utility in terms of individual payoffs will therefore not be considered directly in this paper. $ \mathcal{U} $ can be rather interpreted as a measure of how good a given coalition is according to our criteria since it favors coalitions with good production to risk ratios.

@@

\subsubsection{Filtering the correlation graph}

The previous section explained that coalitions with small variances in their production probability distributions are more likely to afford better contracts. Furthermore, the gaussian approximation highlighted that the correlation structure of the agents plays an important role in finding stable coalitions. Usually, this correlation structure is formalized with a covariance matrix or a correlation matrix that contains all the correlation coefficients between the agents : $ M = (\rho_{ij})_{\forall i,j \in \mathcal{A}^{2}}$.

@@

Selecting the right filter $ \epsilon $ for the $ \epsilon $-graph is an important point since it affects the correlation structure which affects in turn the formation of the coalitions. Unfortunately, there seems to be no clear consensus in the litterature on how to select such a threshold. We will see later in this section that cliques in $ G_{2} $ are potential seeds for the coalitions. Since we want to generate $ N_{COAL} $ coalitions, we need at least $ N_{COAL} $ cliques of a given size to start. Besides, since we consider coalitions as disjoint, the starting cliques should be non overlapping. We thus select our optimal threshold for $ G_{2} $ as :

@@

Where $ G_{2}^{\epsilon} $ is the de-correlation graph $ G_{2} $ filtered by $ \epsilon $, and $ \Theta_{k}(G) $ is the set of non overlapping cliques of size k in a given graph G. In other words we select $ \epsilon^{\star} $ as the smallest threshold posible such that the filtered de-correlation graph contains at least $ N_{COAL} $ non overlapping cliques of size k. The existence of $ \epsilon^{\star} $  as defined in equation  \ref{epsilon_star} is not guaranteed. The users has indeed to provide consistent values of $ N_{COAL} $ or k compared to the size of the agent population $ \mathcal{A} $. \\

\subsubsection{Cliques as de-correlated starting seeds}

Correlation can be seen as cosine of angles in $ L^{2} $, hence even if there is no strict transivity relation for correlation, there is, to a certain extent, some partial notion of it. More precisely, if a, b, and c are three items such that $ \rho_{ab} > \delta $ and $ \rho_{bc} > \delta $, then we know, by the cosine addition formula\footnote{ $ cos(a+b) = cos(a)cos(b) - sin(a)sin(b) $ }, that $ \rho_{ac} > 2 \delta^{2} - 1 $. That is, if a and b are strongly correlated ($\rho_{ab} > 0.9 $) and b and c are also strongly correlated ($\rho_{bc} > 0.9 $), then there is a high probability for a and c of being strongly correlated ($\rho_{ac} > 0.62 $).

Nevertheless de-correlation seems like a more complex concept than correlation in the sense that there is not even a partial notion of transitivity when it comes to it. Therefore, the clustering coefficients of $ G_{1}^{\epsilon} $ is much higher than the one of $ G_{2}^{\epsilon} $. This can be seen as another formulation of Onnela's study on the structural roles of weak and strong links on correlation graphs. Strong links, accounting for strong correlation relationships, are responsible for the clustering, while weak links provide the connectivity between clusters. Searching for clusters in $ G_{2}^{\epsilon} $ and hoping that this strategy will provide a nice coalition structure of internally uncorrelated coalitions seems thus pointless.

Consider now a clique in  $ G_{2}^{\epsilon} $, which is a complete subgraph of $ G_{2}^{\epsilon} $. This is indeed a structure of interest for our purpose. Since there is a link for every pairs of nodes, we know, by construction, that a clique has a mean correlation and a maximum correlation less than $ \epsilon $.

@@

The algorithm presented in the previous section is supposed to generate a given number of coalitions that have good utilities. Nevertheless, as it comprises mainly of a greedy optimization based on local improvements, the probability that the algorithm finds the optimal coalitions set is very low. Actually, it is not obvious that a strict optimum exists at all. Besides, there is, to our knowledge, no state of the art algorithm that aggregate uncorrelated agents in an optimum way (see section \ref{sec:related} for related problems in finance).

In order to have an idea about the algorithm's quality, we compare its results with :
\begin{itemize}
\item \textbf{Random :} This algorithm is basically a random search over the coalition structures space. It analyses a given number of structures and returns the best that it has encountered so far. See algorithm \ref{alg:algo2}.
\item \textbf{Correlated :} This is the complete opposite of our algorithm. It uses the correlation graph $ G_{1} $ and performs a community detection. The resulting coalitions have thus very high internal correlations. We thus expect this algorithm to perform very bad compared to the others. See algorithm \ref{alg:algo3}.
\end{itemize}

Before running the algorithm, we need to calibrate the utility function by choosing the $ \alpha $ parameter. Recall that the purpose of this parameter is to take into account some constraints on the coalition's sizes if needed. In this paper, neither the communication network nor the electrical grid are explicitely considered. Thus, we do not have any technical constraints on coalition sizes even if we designed the utitility such that these could be taken into account. We select the desired size as being $ \lfloor N/N_{COAL} \rfloor $ (where $ \lfloor.\rfloor $ means floor). Figure \ref{fig:real_utility2} shows how the mean utility of a coalition evolves with its size when the optimum size is set to 40 agents. Using equation \ref{eq:alpha_star} to select $ \alpha $ based on the mean quantities seems to give acceptable results for the utility function.

Figure \ref{fig:search} displays the evolution of the global utility and the number of involved agents during the course of the greedy algorithm \ref{alg:algo1}. The transition from unvalid to valid coalitions is clearly visible on the blue diamond curve and occurs between iteration 10 and 15. After this transition, coalition's utilities improve slowly up to a maximum point.

@@

\begin{figure}
\includegraphics[scale=.6]{search.pdf}
\caption{Evolution of the global utility $ \mathcal{U}(CS) $ (blue diamond curve, left axis) and the number of agents involved in the coalitions (red circle curve, right axis) during the greedy optimization of algorithm \ref{alg:algo1} }
\label{fig:search}
\end{figure}
